\vspace{-2.0cm}The effect of nitrogen application time on infection and development of the Wheat Streak Mosaic Virus (WSMV)

### 2015-10-21 leslieandkenny.Rnw
added to comparison section

We address the following questions in this report:
\begin{itemize}
\item Is there evidence that the interaction between nitrogen application and variety differs between inoculated and control plots?
\item Is there evidence that the nitrogen application effect varies across varieties? Is there evidence that the nitrogen application effect differs between inoculated and control?
\item Is there evidence that the variety effect differs between inoculated and control?
\item Conditional on the results to the above questions, how do we appropriately describe the estimated effects for variety, nitrogen application time, and inoculation on the probability of virus infection?
\end{itemize}

@@

\noindent There are several reasons why we recommend using a fixed-effects
generalized linear model instead of a generalized linear mixed model;
these reasons are related to overdispersion and computational issues. \\

\noindent Since the inoculated plants were planted in the center of the
plots, we expect to see spatial dependence where plants near the centers
are more likely to become infected than plants further from the centers.
Then individual plants could have infection probabilities that differ from
the infection rate estimated for the plot they are in. This would result
in the data could having more variation than what the Binomial model
accounts for, a situation known as overdispersion. We observed
overdispersion in Nar's data, and there are no simple tools
for including overdispersion in mixed models. Therefore, the standard
errors in the mixed model output would be based on an incorrect estimate
of the variance and might not be trustworthy. This problem is easily
remedied by using a Quasibinomial fixed-effects GLM model. \\

\noindent Additionally, including many coefficients and nested effects
in a mixed model can cause computational problems where the software will
fail to find unique solutions for the parameters. These problems are made
worse in a Binomial GLMM when the counts are small, as in these data. The
Binomial GLM can avoid many of the computational difficulties associated
with the GLMM. The GLM can correctly model the nested design through a
block by variety interaction. \\

\noindent The only advantage of the GLMM is that it would estimate block-
and row-level variances, but it is our opinion that this is outweighed by
the lack of an overdispersion estimate and the computational difficulties.

@@

\subsection{Comparison of the GLM and the GLMM}

@@

\subsection{Fitting the GLMM}

@@

\noindent For example, if we drop the $3$ way interaction term, the model with all the two-way interactions is fit and the Anova table is considered.

@@

<<refine, echo=FALSE, results='asis'>>=
print(xtable(Anova(glm3way, test="F"), caption="Anova table for the $3$ way interaction model."), caption.placement = "top")
@

@@

<<refitno3way, echo=FALSE, results='asis'>>=
print(xtable(Anova(glm2way, test="F"), caption="Anova table for the model with all $2$ way interactions."), caption.placement = "top")
@

\noindent Here is how we would interpret the results in the rows of interest in the above tables. Note, it is not appropriate to consider removal of any of the terms involving block. Block to block variability is expected because it was included as a design variable, and so it should be included in the model. The $block*variety$ interaction also needs to be included to account for the nesting structure in the design.
\begin{enumerate}
\item \verb+variety:n.trt:status2+ ``There is no evidence that the interaction between variety and nitrogen application time differs between inoculated and control plots (p-value$=0.388$ from F-stat$=1.07$ on $8$ and $125$ df).''
\item \verb+n.trt:status2+ ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection differs between inoculated and control plots after controlling for all other two way interactions and main effects in the model (p-value$=0.380$ from F-stat$=0.973$ on $2$ and $133$ df).
\item \verb+variety:status2+ ``There is no evidence that the relationship between variety and the odds of virus infection differs between inoculated and control plots after controlling for all other two way interactions and main effects in the model (p-value$=0.304$ from F-stat$=1.22$ on $4$ and $133$ df).
\item \verb+variety:n.trt+ ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection depends on variety after controlling for all other two way interactions and main effects in the model (p-value$=0.478$ from F-stat$=0.950$ on $8$ and $133$ df).
\end{enumerate}

\noindent If you remove another term from the model, refit the simpler model and reassess evidence for each effect. For example, suppose we remove the variety by inoculation status interaction term. Our assessment of the variety by nitrogen application interaction effect in the following simpler model would be, ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection depends on variety after controlling for all main effects and the two way interaction between variety and inoculation status (p-value$=0.4536$ from F-stat$=0.98$ on $8$ and $135$ df).

<<refitno2way, echo=FALSE, results='asis'>>=
glm2way2 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt
               + variety*status2,
               family = quasibinomial, data = fert.sim1)
print(xtable(Anova(glm2way2, test="F"), caption="Anova table for the model with two way interactions between variety and nitrogen and variety and inoculation status."), caption.placement = "top")
@

<<refitno2way3, echo=FALSE, results='asis', include=FALSE>>=
glm2way3 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt,
               family = quasibinomial, data = fert.sim1)
print(xtable(Anova(glm2way3, test="F"), caption="Anova table for the model with a two way interaction between variety and nitrogen application time."), caption.placement = "top")
@

\noindent If no evidence is found for any of the interactions, the simplest model we recommend fitting should contain the $block*variety$ interaction and all main effects (shown below). If this model is selected, each effect is then interpreted after controlling for the effects of the other variables in the model. For example, to assess evidence for the \verb+n.trt+ effect in this model, we would say, ``There is no evidence that the odds of virus infection depends on nitrogen application time after controlling for block, variety, inoculation status, and the
$block*variety$ interaction.''

@@

<<glmnointeractions2, echo=FALSE, results='asis'>>=
print(xtable(Anova(glmnointeractions, test="F"), caption="Anova table for the model with no interactions."), caption.placement="top")
@

>>>>

We address the following questions in this report. The first three questions are related to model selection, and the last question is related to interpretation after a model is selected.
\begin{itemize}
\item Is there evidence that the interaction between nitrogen application and variety differs between inoculated and control plots?
\item Is there evidence that the nitrogen application effect varies across varieties? Is there evidence that the nitrogen application effect differs between inoculated and control?
\item Is there evidence that the variety effect differs between inoculated and control?
\item Conditional on the results to the above questions, how do we appropriately describe the estimated effects for variety, nitrogen application time, and inoculation on the probability of virus infection?
\end{itemize}

@@

@@

\subsection{Advantage of the Quasibinomial Generalized Linear Model over the Binomial Generalized Linear Mixed Model}

\noindent There are several reasons why we recommend using a fixed-effects
generalized linear model instead of a generalized linear mixed model in this report. These reasons are related to overdispersion and computational issues. \\

\noindent Since the inoculated plants were planted in the center of the
plots, we expect to see spatial dependence where plants near the centers
are more likely to become infected than plants further from the center. 
Then individual plants could have infection probabilities that differ from
the infection rate estimated for the plot they are in. This would result
in the data could having more variation than what the Binomial model
accounts for, a situation known as overdispersion. We observed
overdispersion in Nar's data, and there are no simple tools
for including overdispersion in mixed models. Therefore, the standard
errors in the mixed model output would be based on an incorrect estimate
of the variance and might not be trustworthy. The standard errors in the GLMM model would be smaller than they should be, causing the p-values for effects of interest to be artificially small. This problem is easily remedied by using a quasibinomial fixed-effects GLM model. \\

\noindent We illustrate this problem in the output below. We fit a quasibinomial logistic regression model, and a binomial generalized linear mixed effects model to the real data sent to us by Nar (details on model fitting in section \ref{model} and \ref{glmmfit}). The coefficients, standard errors, and tests are shown for the three way interaction coefficients in each model. The estimates are similar between the two models, but the standard errors are much larger in the quasibinomial GLM model. Without accounting for overdispersion, the standard errors for the estimates in the mixed effects model are artificially small. For this reason, we advise Nar not to trust the results from the mixed effects model.

<<comparison, echo=FALSE, message=FALSE, results='asis'>>=
glmQBin.noSNMS <- glm(cbind(infected, total) ~ block*variety + 
                        variety*n.trt*status2,
               family = quasibinomial, data = subset(fert2, variety != "SNMS"))
print(xtable(summary(glmQBin.noSNMS)$coefficients[39:44,], caption="Estimates, standard errors, and tests for three way interaction coefficients in the quasibinomial model.", label="qbin"), caption.placement="top", size="footnotesize", table.placement = getOption("xtable.table.placement", "H"))

noSNMS3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
                    (1|block/row), subset = variety != "SNMS",
                    control = glmerControl(optimizer = "bobyqa"),
                    family = binomial, data = fert2)

print(xtable(summary(noSNMS3way)$coefficients[19:24,], caption="Estimates, standard errors, and tests for three way interaction coefficients in the GLMER model", label="glmer"), caption.placement="top", size="footnotesize", table.placement = getOption("xtable.table.placement", "H"))
@

\noindent Additionally, including many coefficients and nested effects
in a mixed model can cause computational problems where the software will
fail to find unique solutions for the parameters. These problems are made
worse in a binomial GLMM when the counts are small, as in these data. The
binomial GLM can avoid many of the computational difficulties associated
with the GLMM.  \\

\noindent Rows are nested within blocks in the study design, and each row is randomly assigned to a variety. The design's nesting structure can be accounted for in a GLMM model  by including random effects for row and block. Originally, however, random effects were developed to reflect a feature of the study design. For example, if $10$ operators were chosen at random from a larger population of operators, {\it operator} would be modeled as a random effect. In this case, the sample of operators truly is random in the design, and the operators selected are intended to represent a larger population of operators. In Nar's study design, row and block are not truly random effects because they were not selected from a larger population of rows and blocks. It is important to recognize this and understand that the only reason row and block are included as random effects in the GLMM model is to account for the nesting structure of row within block. As a result, we think it is more appropriate to model the nested design using fixed effects in the binomial generalized linear model, and this strategy is more consistent with the study design. \\

\noindent The binomial GLM can correctly model the nested design through a block by variety interaction. In the GLM model, it is possible to include a \verb+block/variety+ term, and this is equivalent to including the \verb+block*variety+ interaction. Including this interaction accounts for the nesting because it allows the differences in varieties to vary by block.  \\

\noindent The main advantages of the GLMM is that it would estimate block-
and row-level variances, and it would require fewer degrees of freedom to account for the nesting structure in the design. It is our opinion that this is outweighed by the lack of an overdispersion estimate and the computational difficulties.

@@

\subsection{Fitting the GLMM}
\label{glmmfit}

@@

\noindent The Anova table for the three-way interaction model is shown below (Table \ref{3way}). This table should only be used to assess evidence for the three-way interaction term. We recommend the following wording,
\begin{quote}
There is no evidence that the interaction between variety and nitrogen application time differs between inoculated and control plots (p-value$=0.388$ from F-stat$=1.07$ on $8$ and $125$ df).
\end{quote}

@@

<<refine, echo=FALSE, results='asis'>>=
print(xtable(Anova(glm3way, test="F"), caption="Anova table for the $3$ way interaction model.", label="3way"), caption.placement = "top", size='footnotesize', table.placement = getOption("xtable.table.placement", "H"))
@

\noindent If you drop the $3$ way interaction term, the model with all the two-way interactions should be fit and the Anova table analysis should be repeated.

@@

<<refitno3way, echo=FALSE, results='asis'>>=
print(xtable(Anova(glm2way, test="F"), caption="Anova table for the model with all $2$ way interactions.", label="twoway"), caption.placement = "top", size='footnotesize', table.placement = getOption("xtable.table.placement", "H"))
@

\noindent Here is how we would interpret the results in the applicable rows in the above the two-way interaction Anova table (Table \ref{twoway}). 
\begin{enumerate}
\item \verb+n.trt:status2+ ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection differs between inoculated and control plots after controlling for all other two way interactions and main effects in the model (p-value$=0.380$ from F-stat$=0.973$ on $2$ and $133$ df).
\item \verb+variety:status2+ ``There is no evidence that the relationship between variety and the odds of virus infection differs between inoculated and control plots after controlling for all other two way interactions and main effects in the model (p-value$=0.304$ from F-stat$=1.22$ on $4$ and $133$ df).
\item \verb+variety:n.trt+ ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection depends on variety after controlling for all other two way interactions and main effects in the model (p-value$=0.478$ from F-stat$=0.950$ on $8$ and $133$ df).
\end{enumerate}

\noindent If you remove another term from the model, refit the simpler model and reassess evidence for each effect. For example, suppose we remove the variety by inoculation status interaction term (Table \ref{remove}). Our assessment of the variety by nitrogen application interaction effect in the following simpler model would be,
\begin{quote}
There is no evidence that the relationship between nitrogen application time and the odds of virus infection depends on variety after controlling for all main effects and the two way interaction between variety and inoculation status (p-value$=0.4536$ from F-stat$=0.98$ on $8$ and $135$ df).
\end{quote}
<<refitno2way, echo=FALSE, results='asis'>>=
glm2way2 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt
               + variety*status2,
               family = quasibinomial, data = fert.sim1)
print(xtable(Anova(glm2way2, test="F"), caption="Anova table for the model with two way interactions between variety and nitrogen and variety and inoculation status.", label="remove"), caption.placement = "top", size='footnotesize', table.placement = getOption("xtable.table.placement", "H"))
@

<<refitno2way3, echo=FALSE, results='asis', include=FALSE>>=
glm2way3 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt,
               family = quasibinomial, data = fert.sim1)
print(xtable(Anova(glm2way3, test="F"), caption="Anova table for the model with a two way interaction between variety and nitrogen application time."), caption.placement = "top", table.placement = getOption("xtable.table.placement", "H"))
@

\noindent If no evidence is found for any of the interactions, the simplest model we recommend fitting should contain the $block*variety$ interaction and all main effects (Table \ref{simple}). Note, it is not appropriate to consider removal of any of the terms involving block. Block to block variability is expected because it was included as a design variable, and so it should be included in the model. The $block*variety$ interaction also needs to be included to account for the nesting structure in the design. \\

\noindent To assess evidence for the \verb+n.trt+ effect in the simplest recommended model, we would say, 
\begin{quote}
There is no evidence that the odds of virus infection depends on nitrogen application time after controlling for block, variety, inoculation status, and the $block*variety$ interaction (p-value$=0.612$ from F-stat$=0.49$ on $2$ and $147$ df).
\end{quote}

@@

<<glmnointeractions2, echo=FALSE, results='asis'>>=
print(xtable(Anova(glmnointeractions, test="F"), caption="Anova table for the model with no interactions.", label="simple"), caption.placement="top", size='footnotesize', table.placement = getOption("xtable.table.placement", "H"))
@
